Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
a very quick pdf of "figure 5" done in omnigraffle.  it might be better to do it in dot instead, for consistency with the other lattices -- or remove it entirely, i'm not sure it's worth the space it takes up.
More XXXes.

Note that the left column, i.e.\ the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.  Its
architecture is not relevant to the way it is translated. XXX?

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.
Its lattice order is shown in figure 5.

    [figure 5:                Variable

                         /      |             \           \
                        /       |              \           \
                       /        |               \           \
      Constant(1) ... Constant(n) ... Constant([1,2,3])  ... Constant(<instance of class A>) ...
    ]
\end{verbatim}

@@

The Flow Object Space records all operations that the bytecode
interpreter "would like" to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later. XXX compile-time computation in the helpers.}

@@

Let $x$, $y$ and $z$ be Variables.  We introduce the rule:

% XXX format this!
$$
\begin{array}{c}
 z = \mathrm{add}(x, y), \; b(x) =
    Int, \; Bool \leq b(y) \leq Int \\ \hline
 b' = b \hbox{\ with\ } (z
    \rightarrow Int) 
\end{array}
$$

which specify that if we see the addition operation applied to
Variables whose current binding is $Int$, a new binding $b'$ can be
produced: it is $b$ except on $z$, where we have $b'(z) = Int$.

>>>>

Note that the left column, i.e.\ the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.
The description of this particular input program is beyond the scope of
the present paper; see [S].

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.
Its lattice order is shown in figure \ref{flowlattice}.

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/flowlattice.pdf}
\caption{The lattice order of the flow object space}
\label{flowlattice}
\end{figure*}

@@

The Flow Object Space records all operations that the bytecode
interpreter "would like" to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later.  Moreover, we rely on this feature to perform
compile-time computations, particularly in the generic system code
helpers, which ask for and compute with the concrete types of the
variables they receive.}

@@

Let $x$, $y$ and $z$ be Variables.  We introduce the rules:
%
$$
\begin{array}{c}
 z = \mathrm{add}(x, y), \; b(x) =
    Int, \; b(y) = Int \\ \hline
 b' = b \hbox{\ with\ } (z
    \rightarrow Int) 
\end{array}
$$
%
$$
\begin{array}{c}
 z = \mathrm{add}(x, y), \; Bool \leq b(x), b(y) \leq
    NonNegInt \\ \hline
 b' = b \hbox{\ with\ } (z
    \rightarrow NonNegInt) 
\end{array}
$$
%
The first rule specifies that if we see the addition operation applied to
Variables whose current binding is $Int$, a new binding $b'$ can be
produced: it is $b$ except on $z$, where we have $b'(z) = Int$.
The second rule specifies that in a similar case, if both arguments are
known to be non-negative, so is the result.  Extra rules control addition
between other combinations of numbers, as well as strings, lists and tuples
(for which addition is concatenation).  Similar sets of rules are
introduced for each operation.
